---
title: Point Estimation - A Comprehensive Review # Title of the document

# Bibliography Details
bibliography: ExamReview.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab8WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---

```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
library(ggplot2)
```

# Introduction and Background

Point estimation is a fundamental skill in the statistician's toolbox. Formally, we define a point estimator $\hat{\theta}(X) = \hat{\theta}(X_1, X_2, \dots, X_n)$ as any function of the sample $X$. We use such functions to obtain point estimates, $\hat{\theta}(x) = \hat{\theta}(x_1, x_2, \dots, x_n)$, which are realized values from a point estimator. One might wonder how this applies to practical statistical analysis.


Often when conducting statistical analysis, we do not have information regarding the entire population of interest. As such, we must rely on the available data to estimate the true population parameters. To do this, we compute statistics, or functions of the observed data such as the sample mean, and use them as our parameter estimates. Because statistics are functions of the sample, they are, by definition, point estimators. 

The two methods for point estimation that we covered in MATH240 are (i) Method of Moments and (ii) Maximum Likelihood Estimation. In short, Method of Moments estimation matches sample moments to population moments, while Maximum Likelihood Estimation finds the parameters that maximize the likelihood of observing the existing data.

To demonstrate each procedure, we will work with data from a study on heart failure survival times from @ahmad2017.

 