---
title: Lab Eleven -- Point Estimation and Resampling # Title of the document

# Bibliography Details
bibliography: Lab11References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
library(knitr)
```

**Tips**

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.
* Make sure to use "enter" to make your code readable, so it doesn't run off the page. I switched the Quarto document to automatically render to .pdf, so you can see what you're submitting by default. You can switch back for the highlighting by moving the html block above the pdf block in the YAML header.
* Don't forget that you can copy-and-paste code when you're doing something similar as we did in class!
* Make sure to plot all computed probabilities. This is good `ggplot` practice AND will help make sure you don't make a mistake when computing the probability.


# Introduction
When performing point estimation, there isn't just one *right* answer. For example, the Method of Moments (MOM) and Maximum Likelihood Estimates (MLEs) aren't always the same.

One question that comes up, is why we calculate the sample variance as
$$S_{n-1}^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i -\bar{X})^2,$$
when it is supposed to be the ``average squared distance from the mean, i.e., 
$$S_n^2 = \frac{1}{n} \sum_{i=1}^n (X_i -\bar{X})^2.$$


# Comparing Estimators

## Initial Simulation
In this lab, we will explore why we use $S_{n-1}^2$ and not $S_n^2$. To do this, we will generate data from four known distributions:

* Poisson$(\lambda=2)$
* Binomial$\left(n=50, p=\frac{5-\sqrt{21}}{10}\right)$
* Normal$(\mu=0, \sigma=\sqrt{2})$
* Exponential$\left(\lambda = \frac{1}{\sqrt{2}}\right).$

**Remark**: Note that I have selected the parameters so that the true variance for all four models is 2.

For each distribution, write a loop that completes the following 1000 times. Put `set.seed(7272)` at the top of your code chunk.

* Generate a sample of size $n$ from the distribution.
* Calculate and save $S_{n-1}^2$ and $S_n^2.$

Start with $n=20$. Is there evidence that $S_{n-1}^2$ is a better estimator than $S_n^2$?

Create a plot or combination of plots (using patchwork) to show the results across all four distributions. Make sure to create a corresponding table that numerically summarizes the results of the simulation. You should include
\begin{align*}
\textrm{Bias} &= \textrm{Average of Estimates} - \textrm{True Value}\\
\textrm{Precision} &= \frac{1}{\textrm{Variance of Estimates}}\\
\textrm{Mean Square Error} &= \textrm{Bias}^2 + \textrm{Variance of Estimates}.
\end{align*}

**Solution:**  
*Poisson:*
```{r}
set.seed(7272)

var.n1 = NA #initialized vector for S_{n-1}^2
var.n = NA #initialized vector for S_{n}^2
n = 20
var.true = 2

for(i in 1:1000){
  x = rpois(n = n, lambda = 2)
  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

# Compare estimators
bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)

# Create numerical table
poisson.comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)

poisson.comparison |>
  knitr::kable(digits = 2, caption = "Poisson Variance Estimator Comparison")

# Create Plots
ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

ggdat |> ggplot() +
  geom_density(aes(x=value, color=estimator))+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle("Distribution of Variance Estimators\n(Poisson, Lambda=2)")
```


*Binomial:*
```{r}
set.seed(7272)

var.n1 = NA #initialized vector for S_{n-1}^2
var.n = NA #initialized vector for S_{n}^2
n = 20
var.true = 2

for(i in 1:1000){
  x = rbinom(n,50,(5-sqrt(21))/10)
  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

# Compare estimators
bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)

# Create numerical table
binom.comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)

binom.comparison |>
  knitr::kable(digits = 2, caption = "Binomial Variance Estimator Comparison")

# Create Plots
ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

ggdat |> ggplot() +
  geom_density(aes(x=value, color=estimator))+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle("Distribution of Variance Estimators\n(Binomial, n=50, p=(5-sqrt(21))/10)")
```

*Normal:*
```{r}
set.seed(7272)

var.n1 = NA #initialized vector for S_{n-1}^2
var.n = NA #initialized vector for S_{n}^2
n = 20
var.true = 2

for(i in 1:1000){
  x = rnorm(n,0,sqrt(2))
  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

# Compare estimators
bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)

# Create numerical table
norm.comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)

norm.comparison |>
  knitr::kable(digits = 2, caption = "Normal Dist. Variance Estimator Comparison")

# Create Plots
ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

ggdat |> ggplot() +
  geom_density(aes(x=value, color=estimator))+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle("Distribution of Variance Estimators\n(Normal, mu=0, p=sqrt(2)")
```


*Exponential:*
```{r}
set.seed(7272)

var.n1 = NA #initialized vector for S_{n-1}^2
var.n = NA #initialized vector for S_{n}^2
n = 20
var.true = 2

for(i in 1:1000){
  x = rexp(n,rate=1/sqrt(2))
  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

# Compare estimators
bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)

# Create numerical table
exp.comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)

exp.comparison |>
  knitr::kable(digits = 2, caption = "Exp. Distr. Variance Estimator Comparison")

# Create Plots
ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

ggdat |> ggplot() +
  geom_density(aes(x=value, color=estimator))+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle("Distribution of Variance Estimators\n(Exponential, lambda = 1/sqrt(2)")
```


## Simulation over various sample sizes
Consider the effect of the sample size have on these results. Provide a graphical and numerical summary of what happens as $n$ increases -- say $n=20$, $n=50$, $n=100$, and $n=500$. Does the result in your initial simulation hold across samples? Or, do things change as the sample size changes? Ensure to set your randomization seed.

**Solution:**  
Note: I used `echo=FALSE` to make my PDF easier to look at whilst working on it (I took a "brute force" approach to this section). Apologies if this is an issue.

*Generalizing the previous work into a function (code hidden):* 
```{r}
#| echo=FALSE

distr.simulator = function(n, var.true, distr.name, title.name){
  set.seed(7272)

  var.n1 = NA
  var.n = NA

  for(i in 1:1000){
  
  x = case_when(
      distr.name == "poisson"  ~ rpois(n, 2),
      distr.name == "binomial" ~ rbinom(n, 50, (5 - sqrt(21))/10),
      distr.name == "normal"  ~ rnorm(n, 0, sqrt(2)),
      distr.name == "exponential"   ~ rexp(n, 1/sqrt(2))
    )

  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)
  
comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)


ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

box.plot = ggdat |> ggplot(aes(x = estimator, 
                               y = value, 
                               fill = estimator)) +
  geom_boxplot(alpha=0.7) +
  geom_hline(yintercept = var.true, col = "green") +
  theme_bw() +
  labs(x = "Estimator",
       y = "Variance Estimate",
       fill = "Estimator")

density.plot = ggdat |> ggplot(aes(x=value, color=estimator)) +
  geom_density()+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle(title.name)

return(list(
    table = comparison,
    box.plot = box.plot,
    density.plot = density.plot
  ))

}
```

*Running the Simulations (code hidden):*
Apologies for the looooong brute force coding below. This was just the most intuitive way for me to generate everything!

**Poisson**
```{r}
#| echo=FALSE

pois20 = distr.simulator(20, 2, "poisson", "Poisson, n = 20")
pois50 = distr.simulator(50, 2, "poisson", "Poisson, n = 50")
pois100 = distr.simulator(100, 2, "poisson", "Poisson, n = 100")
pois500 = distr.simulator(500, 2, "poisson", "Poisson, n = 500")

poisson.table = tibble(
  n = c(20, 50, 100, 500),
  `Bias (n-1)` = c(pois20$table$Bias[1], pois50$table$Bias[1], 
                   pois100$table$Bias[1], pois500$table$Bias[1]
  ),
  `Bias (n)` = c(pois20$table$Bias[2], pois50$table$Bias[2], 
                 pois100$table$Bias[2], pois500$table$Bias[2]
  ),               
  `Precision (n-1)` = c(pois20$table$Precision[1], pois50$table$Precision[1], 
                        pois100$table$Precision[1], pois500$table$Precision[1]
  ),                      
  `Precision (n)` = c(pois20$table$Precision[2],      
                      pois50$table$Precision[2], 
                      pois100$table$Precision[2], pois500$table$Precision[2]
  ),
  `MSE (n-1)` = c(pois20$table$MSE[1], pois50$table$MSE[1], 
                  pois100$table$MSE[1], pois500$table$MSE[1]
  ),                
  `MSE (n)` = c(pois20$table$MSE[2], pois50$table$MSE[2], 
                pois100$table$MSE[2], pois500$table$MSE[2]
  )
)


poisson.table |>
  knitr::kable(digits = 4,
               caption = "Poisson Variance Estimators Across Sample Sizes")

# Plots (didn't use patchwork because they were too small)
pois20$box.plot
pois50$box.plot
pois100$box.plot
pois500$box.plot
```

**Binomial**
```{r}
#| echo=FALSE
binom20 = distr.simulator(20, 2, "binomial", "Binomial, n = 20")
binom50 = distr.simulator(50, 2, "binomial", "Binomial, n = 50")
binom100 = distr.simulator(100, 2, "binomial", "Binomial, n = 100")
binom500 = distr.simulator(500, 2, "binomial", "Binomial, n = 500")

binom.table = tibble(
  n = c(20, 50, 100, 500),
  `Bias (n-1)` = c(binom20$table$Bias[1], binom50$table$Bias[1], 
                   binom100$table$Bias[1], binom500$table$Bias[1]
  ),
  `Bias (n)` = c(binom20$table$Bias[2], binom50$table$Bias[2], 
                 binom100$table$Bias[2], binom500$table$Bias[2]
  ),               
  `Precision (n-1)` = c(binom20$table$Precision[1], binom50$table$Precision[1], 
                        binom100$table$Precision[1], binom500$table$Precision[1]
  ),                      
  `Precision (n)` = c(binom20$table$Precision[2],      
                      binom50$table$Precision[2], 
                      binom100$table$Precision[2], binom500$table$Precision[2]
  ),
  `MSE (n-1)` = c(binom20$table$MSE[1], binom50$table$MSE[1], 
                  binom100$table$MSE[1], binom500$table$MSE[1]
  ),                
  `MSE (n)` = c(binom20$table$MSE[2], binom50$table$MSE[2], 
                binom100$table$MSE[2], binom500$table$MSE[2]
  )
)


binom.table |>
  knitr::kable(digits = 4,
               caption = "Binomial Variance Estimators Across Sample Sizes")

# Plots (didn't use patchwork because they were too small)
binom20$box.plot
binom50$box.plot
binom100$box.plot
binom500$box.plot
```


**Normal**
```{r}
#| echo=FALSE
norm20 = distr.simulator(20, 2, "normal", "Normal, n = 20")
norm50 = distr.simulator(50, 2, "normal", "Normal, n = 50")
norm100 = distr.simulator(100, 2, "normal", "Normal, n = 100")
norm500 = distr.simulator(500, 2, "normal", "Normal, n = 500")

norm.table = tibble(
  n = c(20, 50, 100, 500),
  `Bias (n-1)` = c(norm20$table$Bias[1], norm50$table$Bias[1], 
                   norm100$table$Bias[1], norm500$table$Bias[1]
  ),
  `Bias (n)` = c(norm20$table$Bias[2], norm50$table$Bias[2], 
                 norm100$table$Bias[2], norm500$table$Bias[2]
  ),               
  `Precision (n-1)` = c(norm20$table$Precision[1], norm50$table$Precision[1], 
                        norm100$table$Precision[1], norm500$table$Precision[1]
  ),                      
  `Precision (n)` = c(norm20$table$Precision[2],      
                      norm50$table$Precision[2], 
                      norm100$table$Precision[2], norm500$table$Precision[2]
  ),
  `MSE (n-1)` = c(norm20$table$MSE[1], norm50$table$MSE[1], 
                  norm100$table$MSE[1], norm500$table$MSE[1]
  ),                
  `MSE (n)` = c(norm20$table$MSE[2], norm50$table$MSE[2], 
                norm100$table$MSE[2], norm500$table$MSE[2]
  )
)


norm.table |>
  knitr::kable(digits = 4,
               caption = "Normal Distribution Variance Estimators Across Sample Sizes")

# Plots (didn't use patchwork because they were too small)
norm20$box.plot
norm50$box.plot
norm100$box.plot
norm500$box.plot
```


**Exponential**
```{r}
#| echo=FALSE
exp20 = distr.simulator(20, 2, "exponential", "Exponential, n = 20")
exp50 = distr.simulator(50, 2, "exponential", "Exponential, n = 50")
exp100 = distr.simulator(100, 2, "exponential", "Exponential, n = 100")
exp500 = distr.simulator(500, 2, "exponential", "Exponential, n = 500")

exp.table = tibble(
  n = c(20, 50, 100, 500),
  `Bias (n-1)` = c(exp20$table$Bias[1], exp50$table$Bias[1], 
                   exp100$table$Bias[1], exp500$table$Bias[1]
  ),
  `Bias (n)` = c(exp20$table$Bias[2], exp50$table$Bias[2], 
                 exp100$table$Bias[2], exp500$table$Bias[2]
  ),               
  `Precision (n-1)` = c(exp20$table$Precision[1], exp50$table$Precision[1], 
                        exp100$table$Precision[1], exp500$table$Precision[1]
  ),                      
  `Precision (n)` = c(exp20$table$Precision[2],      
                      exp50$table$Precision[2], 
                      exp100$table$Precision[2], exp500$table$Precision[2]
  ),
  `MSE (n-1)` = c(exp20$table$MSE[1], exp50$table$MSE[1], 
                  exp100$table$MSE[1], exp500$table$MSE[1]
  ),                
  `MSE (n)` = c(exp20$table$MSE[2], exp50$table$MSE[2], 
                exp100$table$MSE[2], exp500$table$MSE[2]
  )
)


exp.table |>
  knitr::kable(digits = 4,
               caption = "Exponential Distribution Variance Estimators Across Sample Sizes")

# Plots (didn't use patchwork because they were too small)
exp20$box.plot
exp50$box.plot
exp100$box.plot
exp500$box.plot
```



**Commentary:**   
Across all four of the distributions, we find that the results change with sample size. Specifically, as $n$ grows,the two density plots for each distribution (one for each variance measure) converge, and it becomes harder to differentiate between the density plot for $s^2_{n-1}$ and the density plot for $s^2_{n}$. This phenomenon is further supported by the numerical summary tables. As n increases, we notice that bias and MSE for both $s^2_{n-1}$ and $s^2_{n}$ approach 0, while precision increases for both.

## Simulation
You have written code to draw a random sample of various sizes from each distribution, and to calculate and store the calculated estimates. Rework your code to produce a density plot of the $s_{n-1}^2$ and $s_{n}^2$. Do you have a guess about their distribution? Ensure to set your randomization seed. 

```{r}
pois20$density.plot
pois50$density.plot
pois100$density.plot
pois500$density.plot

binom20$density.plot
binom50$density.plot
binom100$density.plot
binom500$density.plot

norm20$density.plot
norm50$density.plot
norm100$density.plot
norm500$density.plot


exp20$density.plot
exp50$density.plot
exp100$density.plot
exp500$density.plot
```

**Commentary**
Density plots for the four distributions with $n=100$, we can obtain a good estimate about the distributions for $s_{n-1}^2$ and $s_{n}^2$. The first thing the density plots demonstrate is that $s_{n-1}^2$ and $s_{n}^2$ have very similar distributions (despite $s_{n}^2$ being slightly biased toward the left), though what distribution this is depends on the distribution of the model, and the sample size, $n$.   

*Small n:*
For small n, the distributions for $s_{n-1}^2$ and $s_{n}^2$ in all distributions are right-skewed.

*Large n:*
With large enough $n(=500)$, $s_{n-1}^2$ and $s_{n}^2$ for the Poisson, Binomial, Normal, and Exponential distributions, all appear to approach a symmetric distribution. However, for the Exponential distribution, $s_{n-1}^2$ and $s_{n}^2$ still appears to follow a noticeably right-skewed distribution even at $n=100$, taking a bit longer than the others to approach this symmetric shape. 

The graphs also illustrate the reduced variance/improved precision as $n$ increases, with observations further from the mean becoming less likely as $n$ increases.


## Bootstrap
Now, take *one* sample from each distribution. Instead of drawing a random sample at each iteration, sample *with replacement* from the one sample. Produce a density plot of the $s_{n-1}^2$ and $s_{n}^2$. Do we get roughly the same information as we do in the full simulation? Try this over several seeds -- what do you notice? Ensure to set your randomization seed.

**NOTE:**   
Turn off `#|echo=FALSE` to see my clunky code.

```{r}
#| echo=FALSE

distr.bootstrap = function(n, var.true, distr.name, title.name){

  var.n1 = NA
  var.n = NA

 x.orig = case_when(
      distr.name == "poisson"  ~ rpois(n, 2),
      distr.name == "binomial" ~ rbinom(n, 50, (5 - sqrt(21))/10),
      distr.name == "normal"  ~ rnorm(n, 0, sqrt(2)),
      distr.name == "exponential"   ~ rexp(n, 1/sqrt(2))
    )

  for(i in 1:1000){
  
  x = sample(x.orig, size = n, replace = TRUE)

  mean.val = sum(x) / n
  sqr.dev = (x-mean.val)^2
  
  var.n1[i] = sum(sqr.dev) / (n-1)
  var.n[i] = sum(sqr.dev) / n
}

bias.n1 = mean(var.n1 - var.true)
bias.n = mean(var.n - var.true)

precision.n1 = 1/var(var.n1)
precision.n = 1/var(var.n)

mse.n1 = bias.n1^2 + var(var.n1)
mse.n = bias.n^2 + var(var.n)
  
comparison = tibble(
  Estimator = c("Variance (n-1)", "Variance (n)"),
  Bias = c(bias.n1, bias.n),
  Precision = c(precision.n1,precision.n),
  MSE = c(mse.n1, mse.n)
)


ggdat = tibble(
  value = c(var.n1, var.n),
  estimator = rep(c("Variance (n-1)", "Variance (n)"), each = 1000)
)

plot = ggdat |> ggplot() +
  geom_density(aes(x=value, color=estimator))+
  geom_vline(xintercept = var.true, col = "red") +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(x = "Variance Estimate", 
       y = "Density",
       color = "Estimator") +
  ggtitle(title.name)

return(list(
    table = comparison,
    plot = plot
  ))
}
```

```{r}
#| echo=FALSE

set.seed(2567)

pois.boot = distr.bootstrap(100, 2, "poisson", "Bootstrap: Poisson, n = 100")
binom.boot = distr.bootstrap(100, 2, "binomial", "Bootstrap: Binomial, n = 100")
norm.boot = distr.bootstrap(100, 2, "normal", "Bootstrap: Normal, n = 100")
exp.boot = distr.bootstrap(100, 2, "exponential", "Bootstrap: Exponential, n = 100")

boot.table = tibble(
  Distribution = c("Poisson", "Binomial", "Normal","Exponential"),
  `Bias (n-1)` = c(pois.boot$table$Bias[1], binom.boot$table$Bias[1], 
                   norm.boot$table$Bias[1], exp.boot$table$Bias[1]),
  `Bias (n)` = c(pois.boot$table$Bias[2], binom.boot$table$Bias[2], 
                   norm.boot$table$Bias[2], exp.boot$table$Bias[2]),
  `Precision (n-1)` = c(pois.boot$table$Precision[1], binom.boot$table$Precision[1],
                        norm.boot$table$Precision[1], exp.boot$table$Precision[1]),
  `Precision (n)` = c(pois.boot$table$Precision[2], binom.boot$table$Precision[2],
                        norm.boot$table$Precision[2], exp.boot$table$Precision[2]),
  `MSE (n-1)` = c(pois.boot$table$MSE[1], binom.boot$table$MSE[1],
                        norm.boot$table$MSE[1], exp.boot$table$MSE[1]),
  `MSE (n)` = c(pois.boot$table$MSE[2], binom.boot$table$MSE[2],
                        norm.boot$table$MSE[2], exp.boot$table$MSE[2])
)

boot.table |>
  knitr::kable(digits = 4,
               caption = "Bootstrap Variance Estimators Across Distributions")

```

```{r}
pois.boot$plot
binom.boot$plot
norm.boot$plot
exp.boot$plot
```

**Commentary:**
With a moderate sample size of $n=100$, we notice that with bootstrap sampling, the distributions for our estimators exhibit more variance and are less predictable. This is especially noticeable when changing seeds. For example, for one seed (234567), we notice the estimators for the Poisson distribution are significantally more biased than the others, whereas the Exponential distribution exhibits this behavior when the seed is (2567). This demonstrates how bootstrapping is highly dependent on the particular random sample (which is what the seed establishes). Despite the increased variance and bias, the distributions still are "somewhat" close to the true value of 2 (though that description is somewhat liberal).

# Executive Summary
Summarize the work you've done here to provide a brief (200-300 word) summary of the simulation and the results using your work in Lab 11. Your summary should be professional, clear, and all claims should be corroborated with statistical support.


**Response:**
In the above lab, we compare two variance point estimators, $S^2_{n-1}$ and $S^2_{n}$ to identify why using the $n-1$ version is standard practice. We compare the two point estimators through three primary measures (listed in order of priority): (1) Bias, (2) Precision, and (3) Mean Squared Error (MSE). That is to say, we first would like to compare each estimator's bias: if both are unbiased, we compare the precision, if both are biased, we compare the MSE.  

To obtain these values, we simulated 1,000 samples for four distributions: Poisson, Binomial, Normal, and Exponential. Each of these distributions is parameterized such that the true variance is 2, allowing for clear and consistent analysis. For each distribution, we also compared numerical and graphical summaries for multiple n values (20, 50, 100, 500).

The primary findings from this approach is that, regardless of distribution, $S^2_{n}$ is consistently biased downward and $S^2_{n-1}$ is consistently unbiased. This difference is more pronounced for small $n$: as $n$ increases for each of these distributions, the biases of the two estimators converge toward 0. Note that because only $S^2_{n-1}$ is unbiased, we do not need to analyze MSE. Furthermore, precision increases for both estimators with $n$, and the distributions for both estimators improve from right-skewed for small $n$ to symmetric/approximately normal for large $n$.  

After obtaining our primary results, we study if outcomes change under a bootstrapping procedure, and if so, how. The predominant result from this analysis was that bootstrap sampling produces significantly more variable estimator distributions that change wildly with the sampling seed. The bootstrap results still loosely depicted the behavior we exhibited before, but with less reliability.

Overall, this exercise confirms the advantage of using $S^2_{n-1}$ over $S^2_{n}$, particularly when obtaining large samples is not feasible.